Self-tunable approximated explicit MPC: Heat exchanger implementation

### 2023-01-26 paper/elsarticle/main.tex
paper: abstract, conclusion, highlights

\begin{abstract}
max 250 words!

\end{abstract}

@@

%%Research highlights
\begin{highlights}
\item Research highlight 1
\item Research highlight 2
\end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

@@

(skor intro?)The benefit in form of lower computational complexity in the control phase comes hand in hand with a drawback. The size of the parametric solution may be so large, that it becomes impractical to utilize for two reasons: (i) memory footprint is higher than the available memory size of the control unit, (ii) the computational time associated with finding the optimal control action is higher than the available time period for control action implementation. Although this control strategy has its challenges, it is still very beneficial for practical usage for its benefits.

@@

Note, the tuning parameter $\rho$ should be updated only when the reference changes. Updating the tuning parameter in the control steps when $\Delta_{\mathrm{ref}} = 0$ would lead to using tuning parameter with zero value.

TODO: je to znamienkovo symetricke

\subsection{Self-tunable technique for systems with asymmetric behavior}
\label{sec:self_tunable_rho_scaling}
TODO: co ak nepriama umera? \\
TODO: remark na nelinearnu interpolaciu \\
TODO: zovseobecnit na viac usekov intervalu? \\
TODO: intro veta sem\\

The suggested technique of tuning is suitable e.g. for systems with non-symmetric behavior, but can be used in any application, where ,,simple" tuning in the whole range of tuning parameter $\rho$ is not sufficient.

@@

In this section, the results of the proposed tuning method are presented. The plant on which the control was implemented is a laboratory liquid-to-liquid plate heat exchanger Armfield Process Plant Trainer PCT23~\cite{pct23}, see Fig.~\ref{fig:HE}. The cold feed as well as heating medium are transported to the heat exchanger by two peristaltic pumps. The flow rate of the feed is constant, while the aim is to track the reference value of its temperature. Therefore, the controlled variable is the feed temperature $T$ at the outlet of the heat exchanger and the manipulated variable is the voltage $U$ corresponding to the power of the heating medium pump. The voltage is set in percentage, while it is constrained from 20\% to 100\%. As the heat exchange is a nonlinear and non-symmetric process~\cite{Liptak}, this heat exchanger represents a suitable candidate for the presented controller tuning strategy.

@@

The constraints are considered in the terms of manipulated variable, i.e.
\begin{eqnarray}
\label{eq:u_const}
	-15\% \le u \le 65\%.
\end{eqnarray}

Note, the variable $u$ represents the manipulated variable in the deviation form. The values of feed temperature and voltage of the heating medium pump corresponding to zero steady state are respectively $T^\mathrm{s}$~=~35 $^{\circ}\mathrm{C}$ and $U^\mathrm{s}$~=~35\%.

The penalty matrices of the problem Eq.~\eqref{eq:mpc_problem} were systematically tuned and the corresponding control was implemented on the laboratory heat exchanger for every MPC setup. The aim was to determine, which penalty matrix will be the tuned one. Based on observations, the most significant effect on the control trajectories had tuning the penalty matrix $Q_\mathrm{y}$, while still preserving a satisfactory control performance, i.e., without steady-state control error and significant oscillations around the setpoint. The boundary values of matrix $Q_\mathrm{y}$ were tuned as $Q_\mathrm{y, L}$ = 100 and $Q_\mathrm{y, U}$ = 1000. The built-in integrator was penalized with $Q_\mathrm{x}$ = 1 and the manipulated variable with $R$ = 10. The prediction horizon $N$ was 20 steps long. The explicit model predictive controllers were constructed in MATLAB R2020b using the Multi-Parametric Toolbox 3~\cite{mpt_conf}. 
The controllers were implemented to track a time-varying reference. For the first 200 seconds, the system was in the steady state. After that the reference changed its value twice upwards and twice downwards. The reference changes also acquired different sizes in order to examine the proposed tuning method as it is dependent on the size of the reference step change. The control results for both control setups are compared in Fig.~\ref{fig:CV_boundaries} for the controlled variable, and in Fig.~\ref{fig:MV_boundaries} for the manipulated variable.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/CV_boundaries}
		\caption{Controlled variable trajectory for two boundary controllers. The solid lines represent the controlled temperature and the dashed line represents the reference value.}
		\label{fig:CV_boundaries}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/MV_boundaries}
		\caption{Manipulated variable trajectory for two boundary controllers. The solid lines represent the voltage and the dashed lines represent the constraints.}
		\label{fig:MV_boundaries}
	\end{center}
\end{figure}

@@

These observations form the base for the strategy of self-tuning the penalty matrix $Q_\mathrm{y}$. The strategy follows the ideas summarized in Section~\ref{sec:self_tunable_new}.

TODO: dopisat strucne princip ladenia podla toho, ako bude spisana teoria v ~\ref{sec:self_tunable_new}. odkazat sa na rovnice atd...

The control results of the selt-tunable technique compared to the boundary controllers can be seen in Fig.~\ref{fig:CV} for the controlled variable, and in Fig.~\ref{fig:MV} for the manipulated variable. It can be seen that the tuned controller combined the benefits of the two boundary controllers. The overshoots and undershoots were reduced, as in the first half of control the penalty matrix $Q_\mathrm{y}$ acquired value from the first half of the penalty interval. When tracking the reference with negative change, the penalty matrix acquired the values from the second half of the interval, i.e., closer to the upper bound $Q_\mathrm{y,U}$. 
The similarity with the boundary controllers can be seen also on the manipulated variable trajectory. Note, the constraints on the input variable were satisfied as they were scaled based on the boundary controllers which are constructed considering the contraints.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/CV}
		\caption{Controlled variable trajectory for two boundary controllers and the tuned one. The solid lines represent the controlled temperature and the dashed line represents the reference value.}
		\label{fig:CV}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/MV}
		\caption{Manipulated variable trajectory for two boundary controllers and the tuned one. The solid lines represent the voltage and the dashed lines represent the constraints.}
		\label{fig:MV}
	\end{center}
\end{figure}

@@

It can be seen in Table~\ref{tab:control_performance}, that real-time tuning of the controller helped to improve two to three criteria when tracking each reference value. The cost for this is the energy associated with control. Although the integral square error, maximal overshoot/undershoot and settling time decreased, the volume of consumed heating medium did not. Nevertheless, the average deterioration in the terms of consumed heating medium is approximately 17\%.

@@

It can be seen that implementing a tunable controller leads to improved control performance in many criteria. Utilizing a controller with a scalable aggressivity according to the operating conditions leads in general to higher accuracy, lower overshoot and faster achieving the reference value. %It should be also noted that the deterioration of every criterium is calculated subject to the best out of \textit{two} other controllers. It follows that the evaluation is strict as in practice only one controller would be available.
Although the volume of the heating medium is not saved in this control scenario, the improved control performance may lead to other benefits. 
TODO: krajsie skomentovat vyhody zlepsenia kvality riadenia

\section{Conclusion}
\label{sec:conclusion}

>>>>

\begin{abstract}
%max 250 words!
The tunable approximated explicit model predictive control (MPC) comes with the benefits of real-time tunability without the necessity of solving the optimization problem online. This paper provides a self-tunable technique that does not require any further tuning or interventions of the control engineer during operation. Based on the current operating conditions, the tuning parameter adjusts itself and scales the control input using linear interpolation. The adjustment of the tuning parameter depends on the reference value, which makes this strategy suitable for reference tracking problems. Furthermore, a novel technique for scaling the tuning parameter is proposed. It provides to exploit different ranges of the tuning parameter assigned to specified operating conditions. The self-tunable explicit MPC was implemented on a laboratory heat exchanger with nonlinear and non-symmetric behavior. To ensure offset-free control, the built-in integrator was included in the explicit MPC. The asymmetric behavior of the plant was compensated by tuning the controller's aggressivity, as the negativity or positivity of reference change was considered in the tuning process. The control performance decreased maximal overshoots/undershoots, integral square error, and settling time compared to control with only one controller. 
\end{abstract}

@@

%%Research highlights
\begin{highlights}
% 3-5 highlights, every one should not exceed 85 characters including spaces
\item Self-tunable control technique does not need any intervention during control
\item Tuning based on the operating conditions compensates the nonlinear plant behavior
\item Implementation of the proposed tuning method on a laboratory heat exchanger 
\item Control performance improved compared to utilizing only one explicit MPC
\end{highlights}

\begin{keyword}
tunable explicit MPC \sep self-tunable technique \sep tuning parameter \sep heat exchanger \sep control performance
%% keywords here, in the form: keyword \sep keyword

@@

(len skopirovany text z jadra - hodil sa mi viac do Introduction, neskor mozno pouzijem) The benefit in form of lower computational complexity in the control phase comes hand in hand with a drawback. The size of the parametric solution may be so large, that it becomes impractical to utilize for two reasons: (i) memory footprint is higher than the available memory size of the control unit, (ii) the computational time associated with finding the optimal control action is higher than the available time period for control action implementation. Although this control strategy has its challenges, it is still very beneficial for practical usage for its benefits.

@@

Note, only absolute value of $\Delta_{\mathrm{ref}}$ and $d_{\max}$ are considered in this procedure to ensure $\rho \ge 0$.

\begin{remark}
	The tuning parameter $\rho$ should be updated only when the reference changes. Updating the tuning parameter in the control steps when $\Delta_{\mathrm{ref}} = 0$ would lead to using tuning parameter $\rho$ with zero value, i.e., the control input would correspond to one boundary controller and would not be scaled.
\end{remark}

\subsection{Self-tunable technique for systems with asymmetric behavior}
\label{sec:self_tunable_rho_scaling}
\noindent TODO: co ak nepriama umera? \\
TODO: remark na nelinearnu interpolaciu \\
TODO: zovseobecnit na viac usekov intervalu? \\

This paper provides further extension of the self-tuning method proposed in~\cite{self_tunable}. The suggested technique of tuning is suitable e.g. for systems with non-symmetric behavior, but can be used in any application, where ,,simple" tuning in the whole range of tuning parameter $\rho$ is not sufficient.

@@

In this section, the results of the proposed tuning method are presented on an experimental implementation. The tuning strategy utilizes tuning parameter based n the size of reference change (Section~\ref{sec:self_tunable_delta_ref}) and the scaling of tuning parameter based on splitting the interval of the parameter and assigning the interval parts to specific operating conditions (Section~\ref{sec:self_tunable_rho_scaling}).

The plant on which the control was implemented is a laboratory liquid-to-liquid plate heat exchanger Armfield Process Plant Trainer PCT23~\cite{pct23}, see Fig.~\ref{fig:HE}. The cold feed as well as heating medium are transported to the heat exchanger by two peristaltic pumps. The flow rate of the feed is constant, while the aim is to track the reference value of its temperature. Therefore, the controlled output is the feed temperature $T$ at the outlet of the heat exchanger and the control input is the voltage $U$ corresponding to the power of the heating medium pump. The voltage is set in percentage, while it is constrained from 20\% to 100\%. As the heat exchange is a nonlinear and non-symmetric process~\cite{Liptak}, this heat exchanger represents a suitable candidate for the presented controller tuning strategy.

@@

The constraints are considered in the terms of control inputs, i.e.
\begin{eqnarray}
\label{eq:u_const}
	-15\% \le u \le 65\%.
\end{eqnarray}

Note, the variable $u$ represents the control inputs in the deviation form. The values of feed temperature and voltage of the heating medium pump corresponding to zero steady state are respectively $T^\mathrm{s}$~=~35 $^{\circ}\mathrm{C}$ and $U^\mathrm{s}$~=~35\%.

The penalty matrices of the problem Eq.~\eqref{eq:mpc_problem} were systematically tuned and the corresponding control was implemented on the laboratory heat exchanger for every MPC setup. The aim was to determine, which penalty matrix will be the tuned one. Based on observations, the most significant effect on the control trajectories had tuning the penalty matrix $Q_\mathrm{y}$, while still preserving a satisfactory control performance, i.e., without steady-state control error and significant oscillations around the setpoint. The boundary values of matrix $Q_\mathrm{y}$ were tuned as $Q_\mathrm{y, L}$ = 100 and $Q_\mathrm{y, U}$ = 1000. The built-in integrator was penalized with $Q_\mathrm{x}$ = 1 and the control input with $R$ = 10. The prediction horizon $N$ was 20 steps long. The explicit model predictive controllers were constructed in MATLAB R2020b using the Multi-Parametric Toolbox 3~\cite{mpt_conf}. 
The controllers were implemented to track a time-varying reference. For the first 200 seconds, the system was in the steady state. After that, the reference changed its value twice upwards and twice downwards. The reference changes also acquired different sizes in order to examine the proposed tuning method as it is dependent on the size of the reference step change. The control results for both control setups are compared in Fig.~\ref{fig:CV_boundaries} for the controlled outputs, and in Fig.~\ref{fig:MV_boundaries} for the control inputs.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/CV_boundaries}
		\caption{Controlled output trajectory for two boundary controllers. The solid lines represent the controlled temperature and the dashed line represents the reference value.}
		\label{fig:CV_boundaries}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/MV_boundaries}
		\caption{Control input trajectory for two boundary controllers. The solid lines represent the voltage and the dashed lines represent the constraints.}
		\label{fig:MV_boundaries}
	\end{center}
\end{figure}

@@

These observations form the base for the strategy of self-tuning the penalty matrix $Q_\mathrm{y}$. The strategy follows the ideas summarized in Section~\ref{sec:self_tunable_new}. Utilizing the boundary controller with the penalty matrix $Q_\mathrm{y,L}$ is preferred when the reference changes upwards. On the contrary, the controller associated with $Q_\mathrm{y,U}$ is preferred in control with negative reference changes. The splitting value of the tuning parameter was chosen in the middle of the interval, i.e., $\rho_{\mathrm{s}} = 0.5$. The last parameter that needed to be set was the maximal possible size of the reference step change $d_{\max}$, which was set to $15^{\circ}\mathrm{C}$.

The control results of the selt-tunable technique compared to the boundary controllers can be seen in Fig.~\ref{fig:CV} for the controlled outputs, and in Fig.~\ref{fig:MV} for the control inputs. It can be seen that the tuned controller combined the benefits of the two boundary controllers. The overshoots and undershoots were reduced, as in the first half of control the penalty matrix $Q_\mathrm{y}$ acquired value from the first half of the penalty interval. When tracking the reference with negative change, the penalty matrix acquired the values from the second half of the interval, i.e., closer to the upper bound $Q_\mathrm{y,U}$. 
The similarity with the boundary controllers can be seen also on the control inputs trajectory. Note, the constraints on the input variable were satisfied as they were scaled using linear interpolation based on the boundary controllers which are constructed considering the contraints.

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/CV}
		\caption{Controlled output trajectory for two boundary controllers and the tuned one. The solid lines represent the controlled temperature and the dashed line represents the reference value.}
		\label{fig:CV}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{images/MV}
		\caption{Control input trajectory for two boundary controllers and the tuned one. The solid lines represent the voltage and the dashed lines represent the constraints.}
		\label{fig:MV}
	\end{center}
\end{figure}

@@

It can be seen in Table~\ref{tab:control_performance}, that real-time tuning of the controller helped to improve two to three criteria when tracking each reference value. The cost for this is the energy associated with control. Although the integral square error, maximal overshoot/undershoot and settling time decreased, the volume of consumed heating medium did not. Nevertheless, the average deterioration in the terms of consumed heating medium is approximately~17\%.

@@

It can be seen that implementing a tunable controller leads to improved control performance in many criteria. Utilizing a controller with a scalable aggressivity according to the operating conditions leads in general to higher accuracy, lower overshoot and faster achieving the reference value. %It should be also noted that the deterioration of every criterium is calculated subject to the best out of \textit{two} other controllers. It follows that the evaluation is strict as in practice only one controller would be available.
Although the volume of the heating medium is not saved in this control scenario, the improved control performance may lead to other benefits.

TODO: krajsie skomentovat vyhody zlepsenia kvality riadenia

\section{Conclusion}
\label{sec:conclusion}
This paper deals with the experimental implementation of tunable approximated explicit model predictive control and provides a technique for self-tuning controller design. Based on the current reference value, the tuning parameter is scaled using linear interpolation. The previously published paper related to the self-tunable explicit MPC suggested tuning based on the distance of the reference value from the steady state. This paper presents a novel idea of tuning based on the size of reference step change. The tuning procedures aim to compensate for the nonlinear behavior of the controlled system. The tuning parameter is updated when the reference changes. It is calculated as the ratio between the size of the reference change and the maximal possible size of the reference change, which is specified before operation. Another novel contribution comes with splitting the interval of tuning parameter into two ranges, while both ranges are assigned to different operating conditions. The proposed method is implemented on a laboratory heat exchanger with nonlinear and non-symmetric behavior. The non-symmetry makes the plant a suitable candidate for splitting the interval of the tuning parameter. The decision criterium is negativity or positivity of reference change. When the reference changed upwards, the control input was tuned in the first part of the interval and approached the boundary controller associated with the lower bound on the penalty matrix. On the contrary, when the reference changed downwards, the control input was tuned to approach the control input from the boundary controller with the upper bound on the penalty matrix. To investigate the control results properly, the control performance was evaluated quantitatively. Compared to control with only one controller, the self-tunable technique helped to decrease the maximal overshoots/undershoots, integral square error, and settling time.
